---
title: "Conditioning model from CWT and escapement"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditioning model from CWT and escapement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Typical run reconstruction uses a historical time series of spawner and recruit data to estimate the productivity of salmon populations. 
Escapement time series may be available but catch may not be identifiable for a population of interest, for example, if the population is a conservation unit, as part of a larger stock complex and catch composition is not identified to individual CUs.

[Walters and Korman (2024)](https://publications.gc.ca/site/eng/9.940685/publication.html) demonstrated an approach for reconstruction if Coded Wire Tag (CWT) data from an indicator hatchery are assumed to be representative to life cycle parameters of a natural population.

The model consists of two components.
First, CWT data informs natural mortality, maturation, and exploitation rate in the marine environment. 
Second, these parameters are then applied to the system of interest and informs the size and productivity of the population that has a time series of total escapement. The number of hatchery releases in the system informs hatchery production, and the difference in total escapement and hatchery production informs natural production.
Both steps are accomplished within a single model fit, which can account for uncertainty among the various data components, and posterior distributions of parameters are obtained by MCMC.

We utilize this approach as a conditioning model to inform stochastic parameters for projections in salmonMSE, although use of this model is not necessary for projections.

# Life cycle parameters

Life cycle parameters are informed by CWT. Mortality rates are parameterized in instantaneous units, which can be converted to survival terms.

Fishing mortality $F$ is separated into a preterminal (PT) component which acts on juvenile fish and a terminal (T) component on mature fish. Separable effects are modeled where the fishing mortality is year-specific and modified by age class by a vulnerability term $v$.

Survival of juvenile CWT to the next age class $a$ at the beginning of year $y$ is calculated after exploitation, maturation $m$, and natural mortality $M$:

$$
N^\textrm{juv,CWT}_{y+1,a+1} = N^\textrm{juv,CWT}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)(1 - m_{y,a})\exp(-M_{y,a})
$$

The CWT return $R$ is the fraction of maturing juveniles after preterminal exploitation

$$
R^\textrm{CWT}_{y,a} = N^\textrm{juv,CWT}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)m_{y,a}
$$

The escapement is the return that survive terminal exploitation

$$
E^\textrm{CWT}_{y,a} = R^\textrm{CWT}_{y,a}\exp(-v^\textrm{T}_aF^\textrm{T}_y)
$$

The catch $C$ is

\begin{align}
C^\textrm{CWT,PT}_{y,a} &= N^\textrm{juv,CWT}_{y,a}(1 - \exp(-v^\textrm{PT}_aF^\textrm{PT}_y))\\
C^\textrm{CWT,T}_{y,a} &= R^\textrm{CWT}_{y,a}(1 - \exp(-v^\textrm{T}_aF^\textrm{T}_y))
\end{align}

# Hatchery and natural production

The model does separate accounting is natural-origin ($\textrm{NO}$) and hatchery-origin ($\textrm{HO}$) fish in the population of interest. 

The abundance of juvenile fish, recruitment, and escapement uses the same exploitation rate, maturity, and natural mortality parameters estimated from the CWT. 

\begin{align}
N^\textrm{juv,NO}_{y+1,a+1} &= N^\textrm{juv,NO}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)(1 - m_{y,a})\exp(-M_{y,a})\\
N^\textrm{juv,HO}_{y+1,a+1} &= N^\textrm{juv,HO}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)(1 - m_{y,a})\exp(-M_{y,a})
\end{align}

For age 1, the age-1 HO fish is obtained from the number of releases and an assumption about survival after release: $N^\textrm{juv,HO}_{y,a=1} = N^\textrm{rel}_y s^\textrm{rel}$. The age-1 NO fish is calculated from the smolt production (see below).

\begin{align}
R^\textrm{NO}_{y,a} &= N^\textrm{juv,NO}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)m_{y,a}\\
R^\textrm{HO}_{y,a} &= N^\textrm{juv,HO}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)m_{y,a}
\end{align}

\begin{align}
E^\textrm{NO}_{y,a} &= R^\textrm{NO}_{y,a}\exp(-v^\textrm{T}_aF^\textrm{T}_y)\\
E^\textrm{HO}_{y,a} &= R^\textrm{HO}_{y,a}\exp(-v^\textrm{T}_aF^\textrm{T}_y)
\end{align}

The number of spawners is calculated from the escapement, the en-route survival rate, and the proportion allowed to spawn. This proportion can be calculated from the ratio of broodtake to observed escapement:

\begin{align}
\textrm{NOS}_{y,a} &= E^\textrm{NO}_{y,a} s^\textrm{enroute} p^\textrm{spawn}_y\\
\textrm{HOS}_{y,a} &= E^\textrm{HO}_{y,a} s^\textrm{enroute} p^\textrm{spawn}_y
\end{align}

The egg production is calculated from the proportion females and fecundity $f$ at age:

$$
\textrm{Egg}_y = p^\textrm{female}\sum_a f_a (\textrm{NOS}_{y,a} + \gamma \times \textrm{HOS}_{y,a})
$$

The smolt production is calculated from a Ricker stock-recruit function:

$$
\textrm{Smolt}_{y+1} = \alpha\times\textrm{Egg}_y \times \exp(-\beta \times
\textrm{Egg}_y) \exp(-\delta_y)
$$

where $\delta_y$ is an annual deviation from the density-dependent function, expressed as a mortality rate.

# Parameter estimation and priors

## Fishing mortality

Year-specific Fishing mortality is parameterized as 

\begin{align}
F^\textrm{PT}_y &= \exp(a^\textrm{PT}) F^\textrm{trend,PT}_y \exp(\omega^\textrm{FPT}_y)\\
F^\textrm{T}_y &= \exp(a^\textrm{T}) F^\textrm{trend,T}_y \exp(\omega^\textrm{FT}_y)\\
\end{align}

where $F^\textrm{trend,PT}$ is a time series of relative exploitation provided by the analyst. The model estimates a scaling coefficient $a$ and annual deviations $\omega$ to estimate fishing mortality.

The prior for the annual deviations is

\begin{align}
\omega^\textrm{FPT}_y &\sim N(0, \sigma_\textrm{FPT}^2)\\
\omega^\textrm{FT}_y &\sim N(0, \sigma_\textrm{FT}^2)
\end{align}

with hyperpriors for $\sigma_\textrm{FPT}^2$ and $\sigma_\textrm{FT}^2$:

\begin{align}
\sigma_\textrm{FPT} &\sim \textrm{Gamma}(2, 5)\\
\sigma_\textrm{FT} &\sim \textrm{Gamma}(2, 5)
\end{align}

## Vulnerability 

Vulnerability are independent terms estimated in logit space, with relatively uninformative normal priors of mean zero and standard deviation of 1.6:

\begin{align}
\textrm{logit}(v^\textrm{PT}_a) &\sim N(0, 1.6^2)\\
\textrm{logit}(v^\textrm{T}_a) &\sim N(0, 1.6^2)
\end{align}

Vulnerability is fixed at zero and one for the age 1 and maximum age ($A$), respectively ($v_1 = 0$ and $v_A = 1$).

When transformed to normal space, the prior density is relatively uniform between 0-1 with low density at the bounds:

```{r}
x <- seq(-5, 5, 0.1)
f_x <- dnorm(x, 0, 1.6)

y <- plogis(x)
g_y <- f_x /(y * (1 - y)) # Prior density with Jacobian transformation

par(mfcol = c(1, 2), mar = c(5, 4, 1, 1))
plot(x, f_x, typ = 'l', xlab = "logit(v)", ylab = "Prior density")
plot(y, g_y, typ = 'l', xlab = expression(v[a]), ylab = "Prior density")
```

## Maturity

Maturity is estimated in logit space as deviations from base parameters provided by the analyst. The prior density function is Gaussian with separate standard deviation $\sigma^m_a$ by age:

$$
\textrm{logit}(m_{y,a}) \sim N(\textrm{logit}(m_a^\textrm{base}), [\sigma^m_a]^2)
$$

with hyperprior $\sigma^m_a \sim \textrm{Gamma}(2, 5)$.

## Natural mortality

Natural mortality is parameterized as

$$
M_{y,a} = 
\begin{cases}
M^\textrm{base}_a + \sum_iX_{i,y}\beta_i+M^\textrm{add} + \varepsilon_y &, a = 1\\
M^\textrm{base}_a + \sum_jX_{j,y}\beta_j &, a = 2, \ldots, A-1
\end{cases}
$$

From base values provided by the analyst, year-specific mortality rates can be estimated from linear combination of environmental covariates $X$ and estimated coefficients $\beta$. Separate covariates are used for age-1 and age-2+ fish. 
For age-1, an additional scalar $M^\textrm{add}$ and annual deviations $\varepsilon_y$ can be estimated.

Gaussion priors are used for $\varepsilon_y$:

$$
\varepsilon_y \sim N(0, [\sigma^M]^2)
$$
with hyperprior $\sigma^M \sim \textrm{Gamma}(2, 5)$.

## Natural production

The Ricker $\alpha$ parameter is $\alpha = \kappa/\phi_0$, where $kappa$ is productivity and $\phi_0$ is the unfished egg-per-smolt corresponding to the one-to-one recruits per spawner line $\phi_0$, with $\phi_0 = \sum_a \ell_a m^\textrm{base}_a f_a p^\textrm{female}$.

$\ell_a$ is the juvenile survival at age:

$$
\ell_a = 
\begin{cases}
1 &, a = 1\\
\ell_{a-1} \exp(-M^\textrm{base}_{a-1}) (1 - m^\textrm{base}_{a-1}) &, a = 2, \ldots, A
\end{cases}
$$

The $\beta$ parameter is $\beta = \log(\kappa)/\phi_0r_0$, where $r_0$ is the unfished smolt production, first estimated from the unfished spawners $S_0$ parameter and then calculated as $r_0 = S_0/\phi'_0$.

$\phi'_0 = \sum_a \ell_a m^\textrm{base}_a p^\textrm{female}$ is the unfished spawners per smolt.
  
A uniform prior is used for $\log(kappa)$, and a lognormal prior is used to estimate $S_0$.

The annual deviations in smolt production are estimated with prior $\delta_y \sim N(0, \sigma_\delta^2)$ and hyperprior $\sigma_\delta \sim \textrm{Gamma}(2, 5)$.

# Likelihoods

The likelihood of the CWT at age uses a Poisson distribution:

\begin{align}
C^\textrm{CWT,PT}_{y,a} &\sim \textrm{Poisson}(\hat{C}^\textrm{CWT,PT}_{y,a})\\
C^\textrm{CWT,T}_{y,a} &\sim \textrm{Poisson}(\hat{C}^\textrm{CWT,T}_{y,a})\\
E^\textrm{CWT}_{y,a} &\sim \textrm{Poisson}(\hat{E}^\textrm{CWT}_{y,a})
\end{align}

where the $\hat{C}$ and $\hat{E}$ symbols denote estimates of the corresponding observed quantities.

The likelihood of the total escapement for the population of interest uses a lognormal distribution:

$$
\log(E_y) \sim N\left(\log\left(\sum_a(E^\textrm{NO}_{y,a} + E^\textrm{HO}_{y,a})\right), [\sigma^E]^2\right)
$$
