---
title: "Conditioning model from CWT and escapement"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditioning model from CWT and escapement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
```

# Introduction

Typical run reconstruction uses a historical time series of spawner and recruit data to estimate the productivity of salmon populations. 
For many small populations of interest, escapement time series may be available but catch may not be identifiable to the local scale, for example, if the population is a conservation unit (CU) as part of a larger stock complex and catch composition is not identified to individual CUs.

[Walters and Korman (2024)](https://publications.gc.ca/site/eng/9.940685/publication.html) demonstrated an approach for reconstruction if Coded Wire Tag (CWT) data from an indicator hatchery are assumed to be representative to life cycle parameters of a natural population.

The model consists of two components.
First, CWT data informs natural mortality, maturation, and exploitation rate in the marine environment. 
Second, these parameters are then applied to the population of interest and informs the size and productivity of the population that has a time series of total escapement. The number of hatchery releases in the system informs hatchery production, and the difference in total escapement and hatchery production informs natural production.
Both steps are accomplished within a single model fit, which can account for uncertainty among the various data components, and posterior distributions of parameters are obtained by MCMC.

We utilize this approach as a conditioning model to inform stochastic parameters for projections in salmonMSE, although use of this model is not necessary to set up an operating model.

## Model fitting

Model fitting is performed in RTMB with `fit_CM()`. 

The posterior can then be sampled with `sample_CM()` through rstan. 

A subset of posterior MCMC draws of parameters to reconstruct the historical population can be imported into the operating model with `CM2SOM()`.
Further modification of the operating model and additional settings can then be added to run the projection.

# Variable definitions

```{r param-cm, echo = FALSE}
param_cm <- read.csv("tables/glossary_conditioning.csv") %>%
  mutate(Name = paste0("$", Name, "$"))
knitr::kable(param_cm)
```

# Life cycle parameters

Life cycle parameters are informed by CWT. Mortality rates are parameterized in instantaneous units, which can be converted to survival terms.

Fishing mortality $F$ is separated into a preterminal (PT) component which acts on juvenile fish and a terminal (T) component on mature fish. Separable effects are modeled where the fishing mortality is year-specific and modified by age class by a vulnerability term $v$.

Survival of juvenile CWT to the next age class $a$ at the beginning of year $y$ is calculated after exploitation, maturation $m$, and natural mortality $M$:

$$
N^\textrm{juv,CWT}_{y+1,a+1} = N^\textrm{juv,CWT}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)(1 - m_{y,a})\exp(-M_{y,a})
$$

The CWT return $R$ is the fraction of maturing juveniles after preterminal exploitation

$$
R^\textrm{CWT}_{y,a} = N^\textrm{juv,CWT}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)m_{y,a}
$$

The escapement is the return that survive terminal exploitation

$$
E^\textrm{CWT}_{y,a} = R^\textrm{CWT}_{y,a}\exp(-v^\textrm{T}_aF^\textrm{T}_y)
$$

The catch $C$ is

\begin{align}
C^\textrm{CWT,PT}_{y,a} &= N^\textrm{juv,CWT}_{y,a}(1 - \exp(-v^\textrm{PT}_aF^\textrm{PT}_y))\\
C^\textrm{CWT,T}_{y,a} &= R^\textrm{CWT}_{y,a}(1 - \exp(-v^\textrm{T}_aF^\textrm{T}_y))
\end{align}

# Hatchery and natural production

The model does separate accounting of natural-origin ($\textrm{NO}$) and hatchery-origin ($\textrm{HO}$) fish in the population of interest. 

The abundance of juvenile fish $N$, return $R$, and escapement $E$ uses the same exploitation rate, maturity, and natural mortality parameters estimated from the CWT. 

\begin{align}
N^\textrm{juv,NO}_{y+1,a+1} &= N^\textrm{juv,NO}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)(1 - m_{y,a})\exp(-M_{y,a})\\
N^\textrm{juv,HO}_{y+1,a+1} &= N^\textrm{juv,HO}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)(1 - m_{y,a})\exp(-M_{y,a})
\end{align}

For age 1, the age-1 HO fish is obtained from the number of releases and an assumption about survival after release: $N^\textrm{juv,HO}_{y,a=1} = N^\textrm{rel}_y s^\textrm{rel}$. 

The age-1 NO fish is equal to the smolt production: $N^\textrm{juv,NO}_{y,a=1} = \textrm{Smolt}_y$

\begin{align}
R^\textrm{NO}_{y,a} &= N^\textrm{juv,NO}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)m_{y,a}\\
R^\textrm{HO}_{y,a} &= N^\textrm{juv,HO}_{y,a}\exp(-v^\textrm{PT}_aF^\textrm{PT}_y)m_{y,a}
\end{align}

\begin{align}
E^\textrm{NO}_{y,a} &= R^\textrm{NO}_{y,a}\exp(-v^\textrm{T}_aF^\textrm{T}_y)\\
E^\textrm{HO}_{y,a} &= R^\textrm{HO}_{y,a}\exp(-v^\textrm{T}_aF^\textrm{T}_y)
\end{align}

The number of spawners is calculated from the escapement, the en-route survival rate, and the proportion allowed to spawn. This proportion can be calculated from the ratio of broodtake to observed escapement (the model is parameterized in this way for numerical stability and it is assumed broodtake is non-selective with respect to origin):

\begin{align}
\textrm{NOS}_{y,a} &= E^\textrm{NO}_{y,a} s^\textrm{enroute} p^\textrm{spawn}_y\\
\textrm{HOS}_{y,a} &= E^\textrm{HO}_{y,a} s^\textrm{enroute} p^\textrm{spawn}_y
\end{align}

The egg production is calculated from the proportion females and fecundity $f$ at age:

$$
\textrm{Egg}_y = p^\textrm{female}\sum_a f_a (\textrm{NOS}_{y,a} + \gamma \times \textrm{HOS}_{y,a})
$$

The smolt production is calculated from a Ricker stock-recruit function:

$$
\textrm{Smolt}_{y+1} = \alpha\times\textrm{Egg}_y \times \exp(-\beta \times
\textrm{Egg}_y) \exp(-\delta_y)
$$

where $\delta_y$ is an annual deviation from the density-dependent function, expressed as an instantaneous mortality rate. Here, smolts are considered to be outmigrating juveniles. Thus, the survival described by the Ricker function occurs over freshwater life stages.

[Walters and Korman (2024)](https://publications.gc.ca/site/eng/9.940685/publication.html) show that the realized egg-smolt mortality rate can be defined as 

$$
M_y^\textrm{egg} = M^\textrm{egg}_\textrm{min} + \beta \times \textrm{Egg}_y + \delta_y
$$ 

where $M^\textrm{egg}_\textrm{min} = -\log(\phi^{-1}) - \log(\kappa)$ is the density-independent egg mortality rate when the egg production approaches zero, $\beta \times \textrm{Egg}_y$ is the density-dependent mortality term, and $\delta_y$ is a year-specific deviation.

The Ricker $\alpha$ parameter is $\alpha = \kappa/\phi$ in units of juvenile-per-egg, where $\kappa$ is productivity (return/spawner) and $\phi$ is the egg-per-juvenile at replacement (corresponding to the one-to-one return per spawner line), with $\phi = \sum_a \ell_a m^\textrm{base}_a f_a p^\textrm{female}$.

$\ell_a$ is the juvenile survival at age:

$$
\ell_a = 
\begin{cases}
1 &, a = 1\\
\ell_{a-1} \exp(-M^\textrm{base}_{a-1}) (1 - m^\textrm{base}_{a-1}) &, a = 2, \ldots, A
\end{cases}
$$

The $\beta$ parameter is $\beta = \log(\kappa)/(\phi \times J^\textrm{rep})$ is in units of reciprocal eggs, where $J^\textrm{rep}$ is the juvenile smolt production at replacement.
The equilibrium spawners at replacement is $S^\textrm{rep}$ and $J^\textrm{rep} = S^\textrm{rep}/\tau$, where $\tau = \sum_a \ell_a m^\textrm{base}_a p^\textrm{female}$ is the equilibrium female spawners per juvenile.

## Realized productivity

Note that the replacement values described above correspond to initial maturity and natural mortality values. 
The realized replacement, along with the realized productivity and spawners at replacement, varies if either maturity and natural mortality are varying in the historical time series because these parameters change the egg production over the lifetime of a juvenile.

To obtain the realized productivity in a particular year, we can calculate the egg-per-juvenile $\phi_y = \sum_a \ell_{y,a} m_{y,a} f_a p^\textrm{female}$ and spawner-per-juvenile $\tau_y = \sum_a \ell_{y,a} m_{y,a} p^\textrm{female}$ from the historical maturity and mortality, where

$$
\ell_{y,a} = 
\begin{cases}
1 &, a = 1\\
\ell_{y,a-1} \exp(-M_{y,a-1}) (1 - m_{y,a-1}) &, a = 2, \ldots, A
\end{cases}
$$

The realized productivity is $\kappa_y = \alpha \phi_y$ with corresponding $S^\textrm{rep}_y = \log(\kappa_y)\times\tau_y/(\beta\phi_y)$ and $E^\textrm{rep}_y = S^\textrm{rep}_y \phi_y / \tau_y$ ($E^\textrm{rep}_y$ is the egg production at replacement).

The realized productivity can be less than one (and $S^\textrm{rep}_y < 0$) in an individual year. If juvenile marine mortality is high or maturity is late, then there are too few returns to replace the population.

# Parameter estimation and priors

## Fishing mortality

Year-specific fishing mortality is parameterized as 

\begin{align}
F^\textrm{PT}_y &= \exp(a^\textrm{PT}) F^\textrm{trend,PT}_y \exp(\omega^\textrm{FPT}_y)\\
F^\textrm{T}_y &= \exp(a^\textrm{T}) F^\textrm{trend,T}_y \exp(\omega^\textrm{FT}_y)\\
\end{align}

where $F^\textrm{trend,PT}$ is a time series of relative exploitation provided by the analyst. The model estimates a scaling coefficient $a$ and annual deviations $\omega$ to estimate fishing mortality.

The prior for the annual deviations is

\begin{align}
\omega^\textrm{FPT}_y &\sim N(0, \sigma_\textrm{FPT}^2)\\
\omega^\textrm{FT}_y &\sim N(0, \sigma_\textrm{FT}^2)
\end{align}

with hyperpriors for $\sigma_\textrm{FPT}^2$ and $\sigma_\textrm{FT}^2$:

\begin{align}
\sigma_\textrm{FPT} &\sim \textrm{Gamma}(2, 5)\\
\sigma_\textrm{FT} &\sim \textrm{Gamma}(2, 5)
\end{align}

## Vulnerability 

Vulnerability are independent terms estimated in logit space, with relatively uninformative normal priors of mean zero and standard deviation of 1.6:

\begin{align}
\textrm{logit}(v^\textrm{PT}_a) &\sim N(0, 1.6^2)\\
\textrm{logit}(v^\textrm{T}_a) &\sim N(0, 1.6^2)
\end{align}

Vulnerability is fixed at zero and one for the age 1 and maximum age ($A$), respectively ($v_1 = 0$ and $v_A = 1$).

When transformed to normal space, the prior density is relatively uniform between 0-1 with low density at the bounds:

```{r}
x <- seq(-5, 5, 0.1)
f_x <- dnorm(x, 0, 1.6)

y <- plogis(x)
g_y <- f_x /(y * (1 - y)) # Prior density with Jacobian transformation

par(mfcol = c(1, 2), mar = c(5, 4, 1, 1))
plot(x, f_x, type = "l", xlab = "logit(v)", ylab = "Prior density")
plot(y, g_y, typ = 'l', xlab = expression(v[a]), ylab = "Prior density")
```

## Maturity

Maturity is estimated in logit space as deviations from base parameters provided by the analyst. The prior density function is Gaussian with separate standard deviation $\sigma^m_a$ by age:

$$
\textrm{logit}(m_{y,a}) \sim N(\textrm{logit}(m_a^\textrm{base}), [\sigma^m_a]^2)
$$

with hyperprior $\sigma^m_a \sim \textrm{Gamma}(2, 5)$.

## Natural mortality

Natural mortality is parameterized as

$$
M_{y,a} = 
\begin{cases}
M^\textrm{base}_a + \sum_iX_{i,y}\beta_i+M^\textrm{add} + \varepsilon_y &, a = 1\\
M^\textrm{base}_a + \sum_jX_{j,y}\beta_j &, a = 2, \ldots, A-1
\end{cases}
$$

From base values provided by the analyst, year-specific mortality rates can be estimated from linear combination of environmental covariates $X$ and estimated coefficients $\beta$. Separate covariates are used for age-1 and age-2+ fish. 

For age-1, an additional scalar $M^\textrm{add}$ and annual deviations $\varepsilon_y$ can be estimated from the base parameter.

Gaussion priors are used for $\varepsilon_y$:

$$
\varepsilon_y \sim N(0, [\sigma^M]^2)
$$
with hyperprior $\sigma^M \sim \textrm{Gamma}(2, 5)$.

## Natural production

A uniform prior is used for $\log(\kappa)$, and a lognormal prior is used to estimate $S_0$.

The annual deviations in smolt production are estimated with prior $\delta_y \sim N(0, \sigma_\delta^2)$ and hyperprior $\sigma_\delta \sim \textrm{Gamma}(2, 5)$.

# Likelihoods

The likelihood of the CWT at age uses a Poisson distribution:

\begin{align}
C^\textrm{CWT,PT}_{y,a} &\sim \textrm{Poisson}(\hat{C}^\textrm{CWT,PT}_{y,a})\\
C^\textrm{CWT,T}_{y,a} &\sim \textrm{Poisson}(\hat{C}^\textrm{CWT,T}_{y,a})\\
E^\textrm{CWT}_{y,a} &\sim \textrm{Poisson}(\hat{E}^\textrm{CWT}_{y,a})
\end{align}

where the $\wedge$ symbol denotes an estimate of the corresponding observed quantity.

The likelihood of the total escapement for the population of interest uses a lognormal distribution:

$$
\log(E_y) \sim N\left(\log\left(\sum_a(E^\textrm{NO}_{y,a} + E^\textrm{HO}_{y,a})\right), [\sigma^E]^2\right)
$$

# Release strategies

The equations above assume that CWT come from a single release strategy, for example, fed fry, smolts, or yearlings (see [James et al. 2021](https://www.marinescience.psf.ca/wp-content/uploads/2022/03/2021PSF-HatcheryReleaseStrategies-BCReview-FinalReport-Web.pdf) for a review). 

The conditioning model can accommodate multiple release strategies. The number of CWT releases are specified by brood year and release strategy, and the model is fitted to CWT catch by year, age, and release strategy. 

Fishing and natural mortality rates are identical by year and age but maturity differs by release strategy. 

The maturity is estimated by year and age for a particular release strategy indexed by $r = 1$:

$$
\textrm{logit}(m_{y,a,r}) \sim N(\textrm{logit}(m_a^\textrm{base}), [\sigma^m_a]^2)
$$

For all other release strategies $r = 2,\dots,n_r$, logit deviations by age (constant with time) are estimated as fixed effects:

$$
\textrm{logit}(m_{y,a,r}) = \textrm{logit}(m_{y,a,1}) + \delta_{a,r}
$$

The user then specifies the release strategy from which the maturity estimates then apply towards the population of interest.
