---
title: "Dynamics equations"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Dynamics equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
```

While salmonMSE reports state variables by year, the model is configured to model a population by successive generations. 
Thus, the dynamics equations are currently organized by life stage.

As model development continues, the dynamics can be stated by stock and year.

# Definitions

*Definition of variable names and the corresponding slots in either the input (SOM) or output (SMSE) objects in salmonMSE.*

```{r param, echo = FALSE}
param <- read.csv("tables/glossary.csv") %>%
  mutate(Name = paste0("$", Name, "$"))
knitr::kable(param)
```

# Natural production

First, we consider natural production in the absence of fitness effects from the hatchery and habitat improvement.

## Fry

From the spawners (NOS and HOS) of generation $g-1$, the corresponding fry production for generation $g$ is calculated as:

\begin{align}
\textrm{Fry}^\textrm{NOS}_g &= \textrm{NOS}_{g-1} \times p^\textrm{female} \times \textrm{Fec}\\
\textrm{Fry}^\textrm{HOS}_g &= \gamma \times \textrm{HOS}_{\textrm{eff}g-1} \times p^\textrm{female} \times \textrm{Fec}
\end{align}

where $\textrm{HOS}_{\textrm{eff},g-1} = \gamma \times \textrm{HOS}_{g-1}$.

## Smolts

A Beverton-Holt stock-recruit relationship is used to calculate the smolt production

\begin{align}
\textrm{Smolt}^\textrm{NOS}_g &= \frac{P_\textrm{smolt} \times 
\textrm{Fry}^\textrm{NOS}_g}{1 + (\textrm{Fry}^\textrm{NOS}_g + \textrm{Fry}^\textrm{HOS}_g) \frac{P_\textrm{smolt}}{C_\textrm{smolt}}}\\
\textrm{Smolt}^\textrm{HOS}_g &= \frac{P_\textrm{smolt} \times 
\textrm{Fry}^\textrm{HOS}_g}{1 + (\textrm{Fry}^\textrm{NOS}_g + \textrm{Fry}^\textrm{HOS}_g) \frac{P_\textrm{smolt}}{C_\textrm{smolt}}}
\end{align}

# Hatchery production

## Broodtake

The target broodtake is calculated as

$$
\textrm{Broodtake} = \dfrac{n^\textrm{yearling}}{s^\textrm{prespawn} p^\textrm{female} \textrm{Fec}^\textrm{brood} s^\textrm{yearling}} + \dfrac{n^\textrm{subyearling}}{s^\textrm{prespawn} p^\textrm{female} \textrm{Fec}^\textrm{brood} s^\textrm{subyearling}}
$$

The natural broodtake is

$$
\textrm{NOB}_{g-1} = \textrm{max}(p^\textrm{NOB}_\textrm{target} \times \textrm{Broodtake}, p^\textrm{NOB}_\textrm{max} \times \textrm{NOR}^\textrm{escapement}_{g-1})
$$

The hatchery broodtake is

$$
\textrm{HOB}_{g-1} = \textrm{min}(\textrm{Broodtake} - \textrm{NOB}_{g-1}, \textrm{HOR}^\textrm{escapement}_{g-1} \times p^\textrm{HOS}_\textrm{removal})
$$

## Smolt releases

From the total broodtake $\textrm{NOB}_{g-1}$ and $\textrm{HOB}_{g-1}$, the smolt releases is calculated as

$$
\textrm{Smolt}^\textrm{Rel}_g = (\textrm{NOB}_{g-1} + \gamma \times \textrm{HOB}_{g-1}) s^\textrm{prespawn}p^\textrm{female} \textrm{Fec}^\textrm{brood}s^\textrm{hat.}
$$

where the average survival of all hatchery from the egg to yearling stage is

$s^\textrm{hat.} = s^\textrm{subyearling} (1 - p^\textrm{yearling}) + s^\textrm{yearling} p^\textrm{yearling}$  and $p^\textrm{yearling} = n^\textrm{yearling}/(n^\textrm{subyearling} + n^\textrm{yearling})$.

# Pre-terminal fishery

Harvest $u^\textrm{PT}$ in the pre-terminal fishery occurs simultaneously along with marine survival ($\textrm{SAR}$, constant base term). Pre-terminal fishery catch of immature NOS of generation $g$ is calculated as

$$
\textrm{Catch}^\textrm{NOS,PT}_g = \dfrac{F^\textrm{PT}_g}{Z^\textrm{PT}_g}(1 - \exp(-Z^\textrm{PT}_g))[\textrm{Smolt}^\textrm{HOS}_g + \textrm{Smolt}^\textrm{NOS}_g]
$$

where $F^\textrm{PT}_g$ is solved such that $u^\textrm{PT} = \frac{F^\textrm{PT}_g}{Z^\textrm{PT}_g}(1 - \exp(-Z^\textrm{PT}_g))$, $Z^\textrm{PT}_g = F^\textrm{PT}_g + M^\textrm{SAR}_g$.

In the absence of fitness effects, $M^\textrm{SAR}_g = -\log(\textrm{SAR})$.

Similarly, the pre-terminal fishery catch of immature HOS in generation $g$ is

$$
\textrm{Catch}^\textrm{HOS,PT}_g = \dfrac{F_g^\textrm{PT}}{Z^\textrm{PT}_g}(1 - \exp(-Z^\textrm{PT}_g))\textrm{Smolt}^\textrm{Rel}_g
$$

# Recruitment

The recruitment is calculated as the survival of smolts after pre-terminal harvest and marine survival:

\begin{align}
\textrm{NOR}_g &= \exp(-Z^\textrm{PT})[\textrm{Smolt}^\textrm{HOS}_g + \textrm{Smolt}^\textrm{NOS}_g]\\
\textrm{HOR}_g &= \exp(-Z^\textrm{PT})\textrm{Smolt}^\textrm{Rel}_g
\end{align}

## Terminal fishery

Catch of the terminal fishery is calculated from the harvest rate $u^T$:

\begin{align}
\textrm{Catch}^\textrm{NOS,T}_g &= u^\textrm{T} \textrm{NOR}_g\\
\textrm{Catch}^\textrm{HOS,T}_g &= u^\textrm{T} \textrm{HOR}_g
\end{align}

# Escapement

The escapement consists of the survivors of the terminal fishery:

\begin{align}
\textrm{NOR}^\textrm{escapement}_g &= (1 - u^\textrm{T}) \textrm{NOR}_g\\
\textrm{HOR}^\textrm{escapement}_g &= (1 - u^\textrm{T}) \textrm{HOR}_g
\end{align}

# Spawners

The abundance of natural origin spawners consists of the escapement reduced by the broodtake. A separate removal term is used for hatchery origin spawners are removed, where the fish removed can be used as broodtake, research purposes, management of population fitness, etc.

\begin{align}
\textrm{NOS}_g &= \textrm{NOR}^\textrm{escapement}_g - \textrm{NOB}_g\\
\textrm{HOS}_g &= \textrm{NOR}^\textrm{escapement}_g (1 - p^\textrm{HOS}_\textrm{removal})
\end{align}

# Habitat improvement

Habitat improvement can improve either the productivity parameter, capacity parameter, or both. 

Base terms $P_\textrm{smolt}$ and $C_\textrm{smolt}$ are used in the historical period of the model, e.g., for conditioning from data.
In the projection, the stock recruit parameters $P_\textrm{smolt}^\prime$ and $C_\textrm{smolt}^\prime$ are updated with the improvement parameters, $I_P$ or $I_C$, as specified in the operating model:

\begin{align}
P_\textrm{smolt}^\prime &= P_\textrm{smolt}I_P\\
C_\textrm{smolt}^\prime &= C_\textrm{smolt}I_C
\end{align}

# Fitness effects on survival

Reproductive success of hatchery fish has been observed to be lower than their natural counterparts, either through genetic or epigenetic causes, and is accounted for in the $\gamma$ parameter (see review in [Withler et al. 2018](https://www.dfo-mpo.gc.ca/csas-sccs/Publications/ResDocs-DocRech/2018/2018_019-eng.html)).

Through genetic and epigenetic factors, survival of hatchery juveniles in the hatchery environment selects for fish with a phenotype best adapted for that environment, and likewise for juveniles spawned in the natural environment. 
Since these traits are heritable, hatchery programs can shift the fitness of the natural progeny away from the optimum for the natural environment towards that of the hatchery environment on an evolutionary time scale, i.e., over a number of generations.

As described in [Ford 2002](https://doi.org/10.1046/j.1523-1739.2002.00257.x) and derived in [Lande 1976](https://doi.org/10.1111/j.1558-5646.1976.tb00911.x), the fitness loss function $W$ for an individual with phenotypic trait value $z$ in a given environment is

$$
W(z) = \exp\left(\dfrac{-(z-\theta)^2}{2\omega^2}\right)
$$

where $\theta$ is the optimum for that environment and $\omega^2$ is the fitness variance.

If the phenotypic trait value $z$ in the population is a random normal variable with mean $\bar{z}$ and variance $\sigma^2$, then the mean fitness of the population in generation $g$ is $\bar{W}(z) = \int W(z) f(z) dz$. The solution is proportional to

$$
\bar{W}(z) \propto \exp\left(\dfrac{-(\bar{z}-\theta)^2}{2(\omega+\sigma)^2}\right)
$$

We want to follow the mean fitness of generation $g$ for the natural environment:

$$
\bar{W}^\textrm{natural}_g = \exp\left(\dfrac{-(\bar{z}^\textrm{natural}_g-\theta^\textrm{natural})^2}{2(\omega+\sigma)^2}\right)
$$

The mean phenotypic value $\bar{z}$ is calculated iteratively, where the change from generation $g-1$ to $g$ is

\begin{align}
\Delta\bar{z} &= \bar{z}_g - \bar{z}_{g-1} = (\bar{z}^\prime_{g-1} - \bar{z}_{g-1})h^2\\
\bar{z}_g &= \bar{z}_{g-1} + (\bar{z}^\prime_{g-1} - \bar{z}_{g-1})h^2\\
\end{align}

where $h^2$ is the heritability of $z$ and $\bar{z}^\prime_{g-1}$ is the trait value after applying the fitness function, defined as:

\begin{align}
\bar{z}^\prime_{g-1} &= \dfrac{1}{\bar{W}_{g-1}}\int W_{g-1}(z)\times zf(z)dz\\
\bar{z}^\prime_{g-1} &= \left(\dfrac{\bar{z}_{g-1}\omega^2 + \theta \sigma^2}{\omega^2 + \sigma^2} - \bar{z}_{g-1}\right)h^2
\end{align}

With a hatchery program, the mean trait value of the progeny in the natural environment is a weighted average between the mean trait value of natural origin spawners in the natural environment (with optimum value $\theta^\textrm{natural}$) and that of hatchery origin spawners in the hatchery environment (with optimum value $\theta^\textrm{hatchery}$):

$$
\bar{z}^\textrm{natural}_g = p^\textrm{NOS}_{g-1} \times \bar{z}^{\prime\textrm{natural}}_{g-1} + (1 - p^\textrm{NOS}_{g-1}) \times \bar{z}^{\prime\textrm{hatchery}}_{g-1}
$$

where $p^\textrm{NOS} = \textrm{NOS}/(\textrm{NOS} + \gamma\times\textrm{HOS})$. The mean trait value in the hatchery environment $\bar{z}^\textrm{hat.}_g$ is a weighted average of the mean trait value of the natural origin broodtake in the natural environment and that of the hatchery origin broodtake in the hatchery environment:

$$
\bar{z}^\textrm{hatchery}_g = p^\textrm{NOB}_{g-1} \times \bar{z}^{\prime\textrm{natural}}_{g-1} + (1 - p^\textrm{NOB}_{g-1}) \times \bar{z}^{\prime\textrm{hatchery}}_{g-1}
$$

where $p^\textrm{NOB} = \textrm{NOB}/(\textrm{NOB} + \textrm{HOB})$.

The fitness variance and phenotype variance are assumed constant between the two environments. The fitness variance is parameterized relative to the variance of the phenotype, i.e., $\omega = \Omega \sigma$ where $\Omega$ is the "selection strength".

Fitness reduces survival in the egg, fry, and smolt life stages:

\begin{align}
\textrm{Fry}^\textrm{NOS}_g &= \textrm{NOS}_{g-1} \times p^\textrm{female} \times \textrm{Fec} \times (W^\textrm{nat.}_g)^{\ell_\textrm{egg}}\\
\textrm{Fry}^\textrm{HOS}_g &= \gamma \times \textrm{HOS}_{g-1} \times p^\textrm{female} \times \textrm{Fec} \times (W^\textrm{nat.}_g)^{\ell_\textrm{egg}}
\end{align}

\begin{align}
\textrm{Smolt}^\textrm{NOS}_g &= \frac{P_\textrm{smolt}^\prime (W^\textrm{nat.}_g)^{\ell_\textrm{fry}} \times 
\textrm{Fry}^\textrm{NOS}_g}{1 + (\textrm{Fry}^\textrm{NOS}_g + \textrm{Fry}^\textrm{HOS}_g) \frac{P_\textrm{smolt}^\prime}{C_\textrm{smolt}^\prime}}\\
\textrm{Smolt}^\textrm{HOS}_g &= \frac{P_\textrm{smolt}^\prime (W^\textrm{nat.}_g)^{\ell_\textrm{fry}} \times 
\textrm{Fry}^\textrm{HOS}_g}{1 + (\textrm{Fry}^\textrm{NOS}_g + \textrm{Fry}^\textrm{HOS}_g) \frac{P_\textrm{smolt}^\prime}{C_\textrm{smolt}^\prime}}
\end{align}

$$
M^\textrm{SAR}_g = -\log(\textrm{SAR} \times (W^\textrm{nat.}_g)^{\ell_\textrm{smolt}})
$$

where $\ell_i$ is the proportion of the fitness loss apportioned among the three life stages, $\sum_i \ell_i = 1$.

PNI (proportionate natural influence) is an approximation of the rate of gene flow from the hatchery to the natural environment:

$$
\textrm{PNI}_g = \dfrac{p^{\textrm{NOB}}_g}{p^{\textrm{NOB}}_g + (1 - p^{\textrm{NOS}}_g)}
$$

Generally, a combination of minimizing hatchery releases, increasing natural broodtake, and reducing spawning of hatchery origin fish maintains high PNI.
